\Huge Conferences in the Era of Expensive Carbon

### 2019-06-21 main.tex
A mostly finished first draft

Our first proposal is simple: Every ACM-sponsored conference should
calculate and publicly report its carbon footprint. Most conferences'
footprints will be dominated by participants air travel, but the accounting
should go beyond this to include air travel to in-person PC meetings and
estimated hotel and food CO2e footprints.

@@

SIGPLAN has been considering how to reduce CO2e usage over the past couple
of years, informed by an accounting of its CO2e expenditures. This
discussion has led the organizers of two of its flagship conferences (POPL
and ICFP) to switch from in-person to on-line program committee meetings,
joining a trend among other SIGs (\bcp{ZZZ} out of \bcp{WWW} report that
their flagship meetings now have virtual PC meetings). It has also prompted
conferences to increase their investments in livestreaming and talk
recording to permit remote participation.

SIGPLAN has also explored how to mitigate a conference's CO2e emissions by
purchasing \emph{carbon offsets}~\cite{CarbonOFfsetReport}.
%
A carbon offset is sold by a vendor who uses the raised funds to finance an
activity that permanently avoids emitting some amount of greenhouse gases.
The veracity and permanence of this CO2e emission avoidance is certified by
a watchdog organization. (Typically, planting trees is not a certifiable
activity, since it's hard to guarantee the trees won't be cut down, but
installing methane capture devices on landfill sites or buying
fuel-efficient stoves to replace open-fire cooking in poor communities can
be.)
%
Many organizations, including companies such as Google, Dell, Microsoft,
General Motors, Delta Airlines, Lyft, and Expedia, numerous universities,
and some academic societies, now buy carbon offsets in order to reduce their
net footprint. ACM conferences can do the same, and the purchases can be
included in the public accounting we are proposing.  In fact, ACM recently
decided that conferences could offer attendees the option to pay for an
offset during registration; the conference then uses the designated funds to
purchase the offset from a qualified vendor.  Some conferences are also
experimenting with corporate sponsorship of bulk offsets for the whole
conference.\footnote{We acknowledge that carbon offsets are a controversial
  tool~\cite{anderson2012inconvenient,
  carbon-offsets-are-not-our-get-out-jail-free-card}.  Besides issues with
  verifiability and permanence, some observers worry that---especially given
  their currently artificially-low cost---they will encourage complacency,
  leading organizations to postpone making the difficult changes that will
  be required to sustainably decarbonize society.  We believe that offsets
  are a short-term measure, to be considered in parallel with, not instead
  of, more difficult cutting.}

@@

Registration data from recent SIGPLAN conferences illustrates
some of the difficulties.  
%
\bcp{Explain the two diagrams.  Discuss how they can be used to draw
  contradictory conclusions.  E.g., one of the many footprint-reduction
  measures that have been discussed in SIGPLAN~\cite{ClimateCommitteReport}
  is locating conferences near their historical ``center of mass.''  Looking
  at the top diagram, one might thus be tempted to locate all conferences in
  the northeast US or western Europe.  But the bottom figure shows that much
  of the participation in each instance of a conference is local; always
  holding a conference at its ``center of mass'' will damage local
  communities everywhere else.  (Maybe this would be a good place to refer
  to~\cite{ClimateCommitteReport}---we can enumerate some of the
  alternatives that one might have hoped the data would help us choose
  between.)}  
%
\bcp{Maybe we can replace the bottom part of the figure by some clearer way
  of visualizing the observation that much of the attendance at each
  conference is local?  E.g., what if, instead of color-coding the different
  sections of the bars by continent, we instead color-coded them by distance
  traveled (in buckets of 500 miles, say)?}

\bcp{Summary: The data shows why these decisions are genuinely hard and
  painful.}

Nevertheless, it seems clear that the status quo is unsustainable, so hard
choices will have to be made. How do we motivate organizers to make them?

This dilemma is a microcosm of the one that all of society faces. To address
it, many policy experts advocate some form of ``carbon pricing'' (either
taxes or cap-and-trade schemes), which imposes a concrete, immediate cost on
greenhouse gas emissions~\cite{carbonprice}. Doing so makes manifest the
hidden environmental cost of CO2e emissions, thus incentivizing
CO2e-reducing changes without specifying exactly which, allowing for
creative and efficient responses. (Continuing the junk-food analogy, some
municipalities such as New York city and Philadelphia have imposed a
per-calorie tax on soft drinks, with some studies finding that doing so
indeed reduced consumption~\cite{sodatax}.)

Ideally, at some point, governments will impose carbon pricing uniformly,
and all CO2e-intensive activities will have to pay it. But ACM can send a
strong message about the importance of this issue by acting now. In
particular, as several companies have shown, even an artificial ``internal
price'' on carbon creates a significant price signal that shapes behavior.
%
Thus, our second (more ambitious) proposal is: ACM should impose a a
surcharge on conferences based on their carbon footprint.  The charge should
be low at the beginning and increase steadily (and predictably) year on
year.
%
Conference organizers can then choose (on the basis of careful analysis of
data and weighing of benefits) how best to balance their budgets---whether
by decreasing emissions, increasing registration fees, soliciting additional
corporate sponsorship, or by other, more creative, means. \mwh{Problem: We showed
  just above that it's hard to figure out what to do. Come back to
  this point here and say that the surcharge will spur bold choices?
  Maybe say that these choices won't necessarily be bad, but could be
  really great?}

What should ACM do with the funds collected from this surcharge?  One can
imagine many good uses, including buying carbon
offsets, supporting ``green'' computing
research\bcp{can we mention some specifics?}, and defraying the costs of
virtualizing conferences (livestreaming, etc.).  \bcp{Expand this paragraph
  a bit.}

@@

>>>>

Our first proposal is simple: {\bf Every ACM-sponsored conference should
calculate and publicly report its carbon footprint.} Most conferences'
footprints will be dominated by participants air travel, but the accounting
should go beyond this to include air travel to in-person PC meetings and
estimated hotel and food CO2e footprints.

@@

SIGPLAN has been considering how to reduce CO2e usage over the past couple
of years, informed by an accounting of its CO2e expenditures. This
discussion has led the organizers of two of its flagship conferences (POPL
and ICFP) to switch from in-person to on-line program committee meetings,
joining a trend among other SIGs, most of which now use virtual PC meetings
for some or all of their flagship conferences. It has also prompted
conferences to increase their investments in livestreaming and video
recording to support remote participation.

SIGPLAN has also explored how to mitigate a conference's CO2e emissions by
purchasing \emph{carbon offsets}~\cite{CarbonOFfsetReport}.
%
A carbon offset is sold by a vendor who uses the raised funds to finance an
activity that permanently avoids emitting some amount of greenhouse gases.
The veracity and permanence of this CO2e emission avoidance is certified by
a watchdog organization. (Typically, planting trees is not a certifiable
activity, since it's hard to guarantee the trees won't be cut down, but
installing methane capture devices on landfill sites or buying
fuel-efficient stoves to replace open-fire cooking in poor communities can
be.)
%
Many organizations, including companies such as Google, Dell, Microsoft,
General Motors, Delta Airlines, Lyft, and Expedia, numerous universities,
and some academic societies, now buy carbon offsets in order to reduce their
net footprint. ACM conferences can do the same, and the purchases can be
included in the public accounting we are proposing.  In fact, ACM recently
decided that conferences could offer attendees the option to pay for an
offset during registration; the conference then uses the designated funds to
purchase the offset from a qualified vendor.  Some conferences are also
experimenting with corporate sponsorship of bulk offsets for the whole
conference.\footnote{We acknowledge that carbon offsets are a somewhat
  controversial 
  tool~\cite{anderson2012inconvenient,
  carbon-offsets-are-not-our-get-out-jail-free-card}.  Besides issues with
  verifiability and permanence, some observers worry that---especially given
  their currently artificially low cost---they will encourage complacency,
  leading organizations to postpone making the difficult changes that will
  be required to sustainably decarbonize society.  We believe that offsets
  should be considered as a useful short-term expedient, to be considered in
  parallel with, not instead of, more difficult cutting.}

@@

The two visualizations of registration data from some recent conferences in
Figure~\ref{fig:confs} illustrate some of the difficulties.  The top diagram
shows an estimated per-participant CO2e footprint for each instance of the
four major SIGPLAN conferences over the past ten years (excluding a few for
which we had trouble getting data), with larger dots representing more
emissions.  Eyeballing this diagram, it appears that carbon-conscious
conference planners should hold all of these conferences either in the
northeast US or in western Europe every year.  But the bottom visualization
tells a different story.  Here, the horizontal colored bar at the bottom
represents the continent on which each conference was held, and each
vertical bar gives a breakdown of the participants in that conference,
colored according to the continent they gave as their home address.  Again,
from a superficial glance at the colors, it is clear that, though a minority
come from far away, the majority of participants in each case are from the
region where the conference is being held---that is, there is a large amount
of local participation.\iflater\bcp{That point could be strengthened by
  adding some numbers (just in the text, not a figure) about what percentage
  of conference attendees attend even one additional instance of the same
  conference...}\fi{} From this, one might conclude that always locating
conferences in the same one or two places would significantly damage the
research community by discouraging participation from other parts of the
world. 
%
%
% \bcp{Explain the two diagrams.  Discuss how they can be used to draw
%   contradictory conclusions.  E.g., one of the many footprint-reduction
%   measures that have been discussed in SIGPLAN~\cite{ClimateCommitteReport}
%   is locating conferences near their historical ``center of mass.''  Looking
%   at the top diagram, one might thus be tempted to locate all conferences in
%   the northeast US or western Europe.  But the bottom figure shows that much
%   of the participation in each instance of a conference is local; always
%   holding a conference at its ``center of mass'' will damage local
%   communities everywhere else.  (Maybe this would be a good place to refer
%   to~\cite{ClimateCommitteReport}---we can enumerate some of the
%   alternatives that one might have hoped the data would help us choose
%   between.)}  
% %
% \bcp{Maybe we can replace the bottom part of the figure by some clearer way
%   of visualizing the observation that much of the attendance at each
%   conference is local?  E.g., what if, instead of color-coding the different
%   sections of the bars by continent, we instead color-coded them by distance
%   traveled (in buckets of 500 miles, say)?}
%
These disparate conclusions suggest that reducing conference emissions will
require genuinely hard painful compromises.  The impulse to avoid or
postpone thinking about it is entirely understandable!

However, the present trajectory of world emissions is unsustainable, and
hard choices will have to be made if ACM is to play its part by reducing its
own emissions. How do we motivate organizers to make them?

This dilemma is a microcosm of the one that all of society faces. To address
it, many policy experts advocate some form of ``carbon pricing'' (either
taxes or cap-and-trade schemes), which imposes a concrete, immediate cost on
emissions~\cite{carbonprice}. Doing so makes manifest the hidden
environmental cost of CO2e emissions, thus incentivizing CO2e-reducing
changes without specifying exactly which, allowing for creative and
efficient responses. (Continuing the junk-food analogy, some municipalities
such as New York city and Philadelphia have imposed a per-calorie tax on
soft drinks, with some studies finding that doing so indeed reduced
consumption~\cite{sodatax}.)

Ideally, at some point, governments will impose carbon pricing uniformly,
and all CO2e-intensive activities will have to pay it. But ACM can send a
strong message about the importance of this issue by acting now. In
particular, as several companies have shown, even an artificial ``internal
price'' on carbon creates a significant price signal that shapes behavior.
%
Thus, our second proposal is that {\bf ACM should impose a a surcharge on
  conferences based on their carbon footprint.}  The charge should be low at
the beginning and increase steadily (and predictably) year on year.
%
Conference organizers can then choose (on the basis of careful analysis of
data and weighing of benefits) how best to balance their budgets---whether
by decreasing per-participant emissions, decreasing (physical)
participation, increasing registration fees, soliciting additional corporate
sponsorship, or other, more creative, means.  \iflater\mwh{Problem: We
  showed just above that it's hard to figure out what to do. Come back to
  this point here and say that the surcharge will spur bold choices?  Maybe
  say that these choices won't necessarily be bad, but could be really
  great?}\fi

What should ACM do with the funds collected from this surcharge?  One can
imagine many good uses, including carbon offsetting, supporting ``green''
computing research\iflater\bcp{can we mention some specifics?}\fi, and
defraying the costs of virtualizing conferences (livestreaming, etc.).
\iflater\bcp{Expand this paragraph a bit.}\fi

@@

% Ultimately, the carbon cost of air travel may get so high, and individual
% carbon budgets so low, that people simply stop flying except on rare
% occasions.
